Pool Model Paper

### 2023-10-19 pool_model.tex
resource comperition and toxin production, insert results

\section{General Mathemetical Formulation}
Consider the system conssisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

\subsection{Resource-dependent lag-phase}

Another way to encorporate the resource dependence is to assume that the bacteria only leave the lag-phase if there are nutrients, i.e., resources. 
Otherwise they stay in the lag-phase.
Then using the model consisted of two bacteria states $L$ and $G$ we can define the system as
\begin{equation}
    \begin{cases}
        \dot{L} = - \lambda R L\\
        \dot{G} = \lambda R L +\alpha R G\\
        \dot{R} =-\frac{\alpha}{N_t} R G.
    \end{cases}
\end{equation}

Turning back to the equations above we can state: $L+G+N_tR=N_t$. For $t=0$ we have $n_0+N_tR_0=N_t$ (given $L_0=n_0$ and $G_0=0$) and from this: $N_tR_0=N_t-n_0$. Using this we eliminate $R$ and arrive at: 
\begin{equation}
    \begin{cases}
        \dot{L} = - \lambda \left(1-\frac{L+G}{N_t}\right)L\\
        \dot{G} = \lambda \left(1-\frac{L+G}{N_t}\right)L +\alpha \left(1-\frac{L+G}{N_t}\right) G
\end{cases}
\end{equation}
The steady state is $L_s+G_s=N_t$. However, it depends on the initial conditions and the parameters what $L_s$ and $G_s$ are. There is no way to determine $L_s$ and $G_s$ besides solving the dynamical equations, i.e., the steady state cannot be determined directly by solving a set of algebraic equations. To see this we determine the nullclines: the first nullcline is $G=N_t-L$ and the second is $G=N_t-L$, i.e., the nullclines coincides. Every point on the line $G=N_t-L$ is a fixed point. 
This can be seen in Fig. \ref{Phaseplot}. 
For $\lambda\ll \alpha$ the lag-pool is not emptied before all resources are consumed. As a consequence the lag-pool is non-zero in steady-state. 
In contrast, for $\lambda\gg \alpha$ the lag-pool is depleted before the resources are consumed, i.e., the lag-pool is zero in steady state.

The fraction $G$ of the exponentially growing cells will go to steady state due to limitations in resources. 
$\lambda$ is the rate by which cells go from the lag phase to the growth phase; $\mu$ is the rate by which cells go from the lag phase to the state in which they die; $\alpha$ is the growth rate of the healthy population; $\beta$ is the rate by which the cells are removed from the $D$ pool.

\cmt{Plots / results}

@@

We explore now the idea that under certain conditions, as sudden change in the environment, part of the population $G$ will enter a lag-phase, i.e., there is a back-flow from $G$ to $L$. 
\begin{eqnarray}
    L &\stackrel{\lambda}{\longrightarrow} & G\\
    L &\stackrel{\mu}{\longrightarrow} & D\\
    G &\stackrel{\alpha}{\longrightarrow} & 2G\\
    G &\stackrel{\gamma(t)}{\longrightarrow} & L\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}
The rate $\gamma(t)$ depends on the rate of change of the environment. Stronger and quicker changes will lead to a higher rate $\gamma$. 
Let the change in the environment be in temperature $T$ (\textbf{Figure \ref{default}}).

\subsubsection{Maxwell type of stress - strain relation}

@@

\begin{eqnarray}
\label{Model_Comp1}
\dot{L}_A &=& - \lambda_A R L_A\\
\dot{L}_B &=& - \lambda_B R L_B \\
\dot{G}_A &=& \lambda_A R L_A +\alpha_A R G_A\\
\label{Model_Comp4}
\dot{G}_B &=& \lambda_B R L_B +\alpha_B R G_B\\
\dot{R} &=&-\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B
\end{eqnarray}
We again rescaled the resource pool with $N_t$ and find
\begin{eqnarray}
\label{Resource}
R &=&1-\frac{L_A+G_A+L_B+G_B}{N_t}\\
R_0&=&1-\frac{n_A+n_B}{N_t}. 
\end{eqnarray}
Inserting Eq. \ref{Resource} into Eqs. \ref{Model_Comp1} - \ref{Model_Comp4} yields:
\begin{eqnarray*}
\dot{L}_A &=& - \lambda_A  L_A + \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right )L_A\\
\dot{L}_B &=& - \lambda_B L_B + \frac{\lambda_B}{N_t}\left(L_A+G_A+L_B+G_B\right )L_B \\
\dot{G}_A &=&  \lambda_A  L_A - \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right )L_A +\alpha_A G_A - \frac{\alpha_A}{N_t}\left(L_A+G_A+L_B+G_B\right )G_A\\
\dot{G}_B &=& \lambda_B L_B - \frac{\lambda_B}{N_t}\left(L_A+G_A+L_B+G_B\right )L_B  +\alpha_B G_B -\frac{\alpha_B}{N_t}\left(L_A+G_A+L_B+G_B\right )G_B
\end{eqnarray*}

@@

For $N_t\gg\alpha_{A/B}$ the equations decouple. 
The null-clines of the system given by Eqs. \ref{Model_Comp1} - \ref{Model_Comp4} coincide and are given by $N_t=L_A+G_A+L_B+G_B$. 
If $\lambda_{A/B}>\alpha_{A/B}$ the lag-pools are completely depleted before the depletion of the resource pool stops the dynamics. 
To simplify the investigation of the systems dynamics we therefore ignore the $L_A$ and $L_B$ dimension and only consider the dynamics on the centre-manifold given by the $G_A\times G_B$ plane. 
The corresponding phase-plot is shown in Fig. \ref{Competition}. 
The advantage in using the pool model in contrast to using Lotka-Volterra directly is that all parameters have a clear meaning and consistent parameter relations arrive naturally. 
And a species can have more than one internal state. 
However, it is instructive to rewrite the system by setting $x_1\equiv G_A$, $x_2\equiv G_B$, and $\mathbf{x}=(x_1,x_2)^T$:

\begin{eqnarray}
\dot{\mathbf{x}}&=&A\mathbf{x}+\frac{1}{2}\mathbf{x}B\mathbf{x}\\
A&=&
\begin{bmatrix}
\alpha_A&0\\
0 &\alpha_B
\end{bmatrix}\\
B&=&
\frac{-2}{N_t}\begin{bmatrix}
\alpha_A&\alpha_A\\
\alpha_B &\alpha_B
\end{bmatrix}
\end{eqnarray}

The pool model, in contrast to the classical Lotka-Volterra approach, allows for far more flexibility. 
Non-linear functional responses can be introduced without any problem (see how section \ref{Gen_Lot_Vol} for an extended discussion).

\cmt{Plots / results}

@@

\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
\dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
\dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
\dot{R} &=&-\frac{\alpha}{N_t} R G\\
\dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
\mathcal{F}_1(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}\\
\mathcal{F}_2(R,G)&=&\kappa G
\end{eqnarray}
Model $\mathcal{F}_1$ results in waste accumulation in steady state $I_1^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{T}_1=\kappa(\dot{L}+\dot{G})$, which yields: $I_1(t)=\kappa(L(t)+G(t)-n_0)$ where we used $T_1(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide. \\

Model $\mathcal{F}_2$ results in $I_2(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ toxin is produced, no matter whether the amount of bacteria is still growing. 
The amount of toxin depends therefore on the back-flow model employed, which can be seen in Fig. \ref{Waste}.

>>>>

\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

@@

We explore now the idea that under certain conditions, as sudden change in the environment, part of the population $G$ will enter a lag-phase, i.e., there is a back-flow from $G$ to $L$. 
\begin{eqnarray}
    L &\stackrel{\lambda}{\longrightarrow} & G\\
    L &\stackrel{\mu}{\longrightarrow} & D\\
    G &\stackrel{\alpha}{\longrightarrow} & 2G\\
    G &\stackrel{\gamma(t)}{\longrightarrow} & L\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}
The rate $\gamma(t)$ depends on the rate of change of the environment.
Stronger and quicker changes will lead to a higher rate $\gamma$. 
\subsubsection{Maxwell type of stress - strain relation}

@@

\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A +\alpha_A R G_A\\
        \dot{G}_B = \lambda_B R L_B +\alpha_B R G_B\\
        \dot{R} =-\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B
    \end{cases}
    \label{eq:model_2sp_resource_comp}
\end{equation}

We again rescaled the resource pool with $N_t$ and find
\begin{eqnarray}
\label{Resource}
R &=&1-\frac{L_A+G_A+L_B+G_B}{N_t}
\end{eqnarray}
Inserting Eq. \ref{Resource} into Eqs. \ref{eq:model_2sp_resource_comp} yields:
\begin{eqnarray*}
\dot{L}_A &=& - \lambda_A  L_A + \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right )L_A\\
\dot{L}_B &=& - \lambda_B L_B + \frac{\lambda_B}{N_t}\left(L_A+G_A+L_B+G_B\right )L_B \\
\dot{G}_A &=&  \lambda_A  L_A - \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right )L_A +\alpha_A G_A - \frac{\alpha_A}{N_t}\left(L_A+G_A+L_B+G_B\right )G_A\\
\dot{G}_B &=& \lambda_B L_B - \frac{\lambda_B}{N_t}\left(L_A+G_A+L_B+G_B\right )L_B  +\alpha_B G_B -\frac{\alpha_B}{N_t}\left(L_A+G_A+L_B+G_B\right )G_B
\end{eqnarray*}

@@

The pool model, in contrast to the classical Lotka-Volterra approach, allows for far more flexibility.

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=0.67\textwidth]{Figures/pool_model_2pools_resource_competition.pdf}
    \caption{{\footnotesize The competition of the two bacterial species due to the common limiting resource calculated using Eq. (\ref{eq:model_2sp_resource_comp}).
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The purple dashed line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_resource_2sp}
    \end{center}
\end{figure}
As one can see in Figure~\ref{fig:2pool_resource_2sp} even without direct interaction between species the bacterial growth was inhibited due to the limitation of the resource.
Here species $A$ was able to supress species $B$ due to the higher values of the growth constant $\alpha$..

@@

\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
\dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
\dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
\dot{R} &=&-\frac{\alpha}{N_t} R G\\
\dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
\mathcal{F}_1(R,G)&=&\kappa G\\
\mathcal{F}_2(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}
\end{eqnarray}
We would like to test the following model:
\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\
        \dot{G}_B = \lambda_B R L_B + \alpha_B R G_B\\
        \dot{R} = -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B\\
        \dot{T} = \mathcal{F}_{1,2} (R, G_B) \\
    \end{cases}
    \label{eq:model_2sp_toxin}
\end{equation}

As it can be seen in Figure~\ref{fig:2pool_2sp_toxin}, model $\mathcal{F}_1$ results in $I_1(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ toxin is produced, no matter whether the amount of bacteria is still growing. 
Model $\mathcal{F}_2$ results in waste accumulation in steady state $I_2^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{T}_2=\kappa(\dot{L}+\dot{G})$, which yields: $I_2(t)=\kappa(L(t)+G(t)-n_0)$ where we used $T_2(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide. \\

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=1.\textwidth]{Figures/pool_model_2pools_toxin.pdf}
    \caption{{\footnotesize The two species system competing for the common nutritient $R$ with production of the toxin $T$ (brown solid line) calculated using Eq. (\ref{eq:model_2sp_toxin}).
    The growing bacteria $B$ produces toxin $T$ (a) constantly $\dot{T} = \mathcal{F}_1 = \kappa G_B$ and (b) only during growth process $\dot{T} = \mathcal{F}_2 = \kappa \alpha_B R G_B$ with 
    production rate $\kappa=0.2$.
    The toxin $T$ causes the cell death of bacteria $A$ with rate constant $\mu = 0.2$.
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_2sp_toxin}
    \end{center}
\end{figure}
